---
title: "IPHC data - rockfish"
author: "Alberto Rovellini"
date: "10/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rbgm)
library(sf)
library(viridis)
library(maps)
library(mapdata)
library(data.table)
```

```{r}
select <- dplyr::select
```

The Slope Assemblage Rockfish species, which include some of the GOA's most valuable like northern, shortraker, rougheye and blackspotted, seemed to be captured okay in the bottom trawl, with models converging both in Alaska and Canada for both juveniles and adults.

Demersal shelf rockfish also converged in both areas, but the AFSC data had few data points for these species. in addition, DSR exists only in SE Alaska, and the species that compose it are managed as other rockfish everywhere else (rockfish management is confusing). Models for pelagic shelf rockfish did not converge in Alaska with 5 knots for the depth spline, but were re-run with 3 knots and converged. PSR are not well sampled by the IPHC surveys (few data points).

So, for DSR and PSR let's have a lott at the IPHC data, and see if we can use that to fill some of the gaps.

Need to look into what this RPN business is.
```{r}
dat <- read.csv('../data/IPHC/fiss_cleaned_09222021.csv')

atlantis_bgm <- read_bgm('../data/GOA_WGS84_V4_final.bgm')
atlantis_box <- atlantis_bgm %>% box_sf()
# utilities
atlantis_crs <- atlantis_bgm$extra$projection
atlantis_bbox <- atlantis_box %>% st_bbox()
```

Check how many data points there are for each species, and which years we have surveys for.
```{r}
dat %>% select(SPECIES_COMMON_NAME) %>% group_by(SPECIES_COMMON_NAME) %>% tally() %>% arrange(desc(n))
dat %>% select(SURVEY_YEAR) %>% distinct() %>% pull() %>% sort()
```

Read in the species key from the RACE data - it should map okay to IPHC too.
```{r}
key <- read.csv('../data/RACE_species_goa_Atlantis_Nov162021.csv')
rfd <- key %>% filter(Atlantis.group=='RFD') %>% pull(Scientific.Name) %>% unique()
#rfp <- key %>% filter(Atlantis.group=='RFP') %>% pull(Scientific.Name) %>% unique()
```

Extract rockfish of the demersal and pelagic shelf assemblages.
```{r}
species_iphc <- dat %>% select(SPECIES_SCIENTIFIC_NAME) %>% distinct() %>% arrange(SPECIES_SCIENTIFIC_NAME) %>%  pull()

rfd_iphc <- intersect(species_iphc,rfd)
#rfp_iphc <- intersect(species_iphc,rfp)

rockfish <- dat %>% # repeat this for demersal and pelagic
  filter(SPECIES_SCIENTIFIC_NAME %in% rfp_iphc & IPHC_REG %in% c('4A','3B','3A','2C','2B')) %>%
  select(SPECIES_SCIENTIFIC_NAME,FISHING_EVENT_ID,SURVEY_YEAR,IPHC_REG,START_LAT,START_LON,AVG_DEPTH_M,HOOKS_RETRIEVED,EX_EFF_HOOKS,SUBSAMPLE,OBS_CATCH) %>%
  group_by(across(FISHING_EVENT_ID:SUBSAMPLE)) %>%
  summarise(OBS_CATCH = sum(OBS_CATCH)) %>%
  ungroup()
```

View in space.
```{r}
coast <- maps::map("worldHires", c("USA","Canada"), plot = FALSE, fill = TRUE)
coast_sf <- coast %>% st_as_sf() #%>% st_transform(crs = atlantis_crs)

rockfish_sf <- rockfish %>% st_as_sf(coords=c('START_LON','START_LAT'),crs=4326)
rockfish_bbox <- rockfish_sf %>% st_bbox()

rockfish_sf %>% ggplot()+
  geom_sf(aes(color=log1p(OBS_CATCH)))+
  scale_color_viridis()+
  geom_sf(data=coast_sf)+
  coord_sf(xlim=c(rockfish_bbox$xmin,rockfish_bbox$xmax), ylim=c(rockfish_bbox$ymin,rockfish_bbox$ymax))+
  facet_wrap(~SURVEY_YEAR)+
  theme_bw()
```

Dusky rockfish not sampled by the IPHC either - which is interesting. And even the demersal shelf are not super well-represented. 

Let's get the fishing events in these Regulatory Areas that did not catch any rockfish.
```{r}
these_events <- dat %>% filter(IPHC_REG %in% c('4A','3B','3A','2C','2B')) %>%
  select(FISHING_EVENT_ID,SURVEY_YEAR,IPHC_REG,START_LAT,START_LON,AVG_DEPTH_M,HOOKS_RETRIEVED,EX_EFF_HOOKS,SUBSAMPLE,OBS_CATCH) %>% 
  select(FISHING_EVENT_ID) %>% 
  distinct() %>% 
  pull()

rockfish_events <- rockfish %>%  select(FISHING_EVENT_ID) %>%  distinct() %>% pull()

# build a dataframe with events that had zero catch for sharks
no_rockfish_events <- setdiff(these_events,rockfish_events)

no_rockfish_frame <- dat %>% 
  filter(FISHING_EVENT_ID %in% no_rockfish_events & IPHC_REG %in% c('4A','3B','3A','2C','2B')) %>% 
  select(FISHING_EVENT_ID,SURVEY_YEAR,IPHC_REG,START_LAT,START_LON,AVG_DEPTH_M,HOOKS_RETRIEVED,EX_EFF_HOOKS,SUBSAMPLE) %>% 
  distinct() %>%
  mutate(OBS_CATCH=0)

rockfish_all <- rbind(rockfish,no_rockfish_frame) # slightly longer than the sum of events, meaning that some events are duplicated. Explore below

dups <- rockfish_all %>% group_by(FISHING_EVENT_ID) %>% tally() %>% filter(n>1) %>% select(FISHING_EVENT_ID) %>% pull()
dups_rockfish <- rockfish_all %>% filter(FISHING_EVENT_ID %in% dups) %>% distinct() # weirdly, it picks up some differences between pairs of these, although they look the same. probably negligible difference
```

View.
```{r}
rockfish_all_sf <- rockfish_all %>% st_as_sf(coords=c('START_LON','START_LAT'),crs=4326)
rockfish_all_bbox <- rockfish_all_sf %>% st_bbox()

rockfish_all_sf %>% ggplot()+
  geom_sf(aes(color=log1p(OBS_CATCH)))+
  scale_color_viridis()+
  geom_sf(data=coast_sf)+
  coord_sf(xlim=c(rockfish_bbox$xmin,rockfish_bbox$xmax), ylim=c(rockfish_bbox$ymin,rockfish_bbox$ymax))+
  facet_wrap(~SURVEY_YEAR)+
  theme_bw()
```
This is a big dataset with a lot of points. May be worth trying with sdmTMB, the issue is getting a CPUE from the hook counts.

Have a look at the observed catch per subsampling. The metadata say about the 'subsample' field: "Sampling scheme (1=20 hooks counted near the beginning of skate is standard in U.S. reg areas except one year in SEAK for yelloweye; 2=100% hook count is standard in Canadian reg areas)". However, the data contain 0 and 1 as values for SUBSAMPLE.
```{r}
subsample <- dat %>% select(OBS_CATCH,SUBSAMPLE,IPHC_REG)

t <- subsample %>% group_by(IPHC_REG,SUBSAMPLE) %>% tally() # area 2B (Canada) seems to be the only area where there is a prevalence of 0, which I'd interpret as (no subsampling)

subsample %>% #filter(OBS_CATCH>0) %>% 
  ggplot()+
  geom_boxplot(aes(x = factor(SUBSAMPLE), y=OBS_CATCH)) # 0 seems to present higher catches, hinting that 0 means no subsampling
```

It seems that if SUBSAMPLE=1 they do the subsampling, i.e. they only count the first 20 hooks. If SUBSAMPLING=0, they count all hooks on the skate. So, if SUBSAMPLING==1, OBS_CATCH=OBS_CATCH*5.
```{r}
rockfish_all <- rockfish_all %>% rowwise() %>% mutate(CATCH=ifelse(SUBSAMPLE==1,OBS_CATCH*5,OBS_CATCH))
```

Not 100% convinced that the adjustment for subsample is correct. Divide by hook number to get NPUE.
```{r}
rockfish_all <- rockfish_all %>% mutate(NPUE=CATCH/EX_EFF_HOOKS)
```

Prepare this for the sdmTMB code.
```{r}
rockfish_all <- rockfish_all %>% select(SURVEY_YEAR,FISHING_EVENT_ID,START_LAT,START_LON,AVG_DEPTH_M,NPUE) %>% 
  set_names(c(
  "year",
  "hauljoin",
  "lat",
  "lon",
  "depth",
  "npue"))

saveRDS(rockfish_all,'../data/IPHC/rockfish_rfp.RDS')
```
