---
title: "IPHC data"
author: "Alberto Rovellini"
date: "10/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rbgm)
library(sf)
library(viridis)
library(maps)
library(mapdata)
library(data.table)
```

```{r}
select <- dplyr::select
```

Explore the IPHC longline data, to get some distributions of sleeper sharks hopefully. 

Need to look into what this RPN business is.
```{r}
dat <- read.csv('../data/IPHC/fiss_cleaned_09222021.csv')

atlantis_bgm <- read_bgm('../data/GOA_WGS84_V4_final.bgm')
atlantis_box <- atlantis_bgm %>% box_sf()
# utilities
atlantis_crs <- atlantis_bgm$extra$projection
atlantis_bbox <- atlantis_box %>% st_bbox()
```

Extract sleeper sharks.
```{r}
#dat %>% select(SPECIES_COMMON_NAME) %>% distinct() %>% arrange(SPECIES_COMMON_NAME) %>%  pull()

sleep <- dat %>% 
  filter(SPECIES_COMMON_NAME=='Sleeper Shark' & IPHC_REG %in% c('4A','3B','3A','2C','2B')) %>%
  select(FISHING_EVENT_ID,SURVEY_YEAR,IPHC_REG,START_LAT,START_LON,AVG_DEPTH_M,HOOKS_RETRIEVED,EX_EFF_HOOKS,SUBSAMPLE,OBS_CATCH)
```

View in space.
```{r}
coast <- maps::map("worldHires", c("USA","Canada"), plot = FALSE, fill = TRUE)
coast_sf <- coast %>% st_as_sf() #%>% st_transform(crs = atlantis_crs)

sleep_sf <- sleep %>% st_as_sf(coords=c('START_LON','START_LAT'),crs=4326)
sleep_bbox <- sleep_sf %>% st_bbox()

sleep_sf %>% ggplot()+
  geom_sf(aes(color=log1p(OBS_CATCH)))+
  scale_color_viridis()+
  geom_sf(data=coast_sf)+
  coord_sf(xlim=c(sleep_bbox$xmin,sleep_bbox$xmax), ylim=c(sleep_bbox$ymin,sleep_bbox$ymax))+
  facet_wrap(~SURVEY_YEAR)+
  theme_bw()
```

Let's get the fishing events in these Regulatory Areas that did not catch any sleeper sharks.
```{r}
these_events <- dat %>% filter(IPHC_REG %in% c('4A','3B','3A','2C','2B')) %>%
  select(FISHING_EVENT_ID,SURVEY_YEAR,IPHC_REG,START_LAT,START_LON,AVG_DEPTH_M,HOOKS_RETRIEVED,EX_EFF_HOOKS,SUBSAMPLE,OBS_CATCH) %>% 
  select(FISHING_EVENT_ID) %>% 
  distinct() %>% 
  pull()

sleep_events <- sleep %>%  select(FISHING_EVENT_ID) %>%  distinct() %>% pull()

# build a dataframe with events that had zero catch for sharks
no_sleep_events <- setdiff(these_events,sleep_events)

no_sleep_frame <- dat %>% 
  filter(FISHING_EVENT_ID %in% no_sleep_events & IPHC_REG %in% c('4A','3B','3A','2C','2B')) %>% 
  select(FISHING_EVENT_ID,SURVEY_YEAR,IPHC_REG,START_LAT,START_LON,AVG_DEPTH_M,HOOKS_RETRIEVED,EX_EFF_HOOKS,SUBSAMPLE) %>% 
  distinct() %>%
  mutate(OBS_CATCH=0)

sleep_all <- rbind(sleep,no_sleep_frame) # slightly longer than the sum of events, meaning that some events are duplicated. Explore below

dups <- sleep_all %>% group_by(FISHING_EVENT_ID) %>% tally() %>% filter(n>1) %>% select(FISHING_EVENT_ID) %>% pull()
dups_sleep <- sleep_all %>% filter(FISHING_EVENT_ID %in% dups) %>% distinct() # weirdly, it picks up some differences between pairs of these, although they look the same. None of these contained sleeper shark catches, so probably negligible difference
```

View.
```{r}
sleep_all_sf <- sleep_all %>% st_as_sf(coords=c('START_LON','START_LAT'),crs=4326)
sleep_all_bbox <- sleep_all_sf %>% st_bbox()

sleep_all_sf %>% ggplot()+
  geom_sf(aes(color=log1p(OBS_CATCH)))+
  scale_color_viridis()+
  geom_sf(data=coast_sf)+
  coord_sf(xlim=c(sleep_bbox$xmin,sleep_bbox$xmax), ylim=c(sleep_bbox$ymin,sleep_bbox$ymax))+
  facet_wrap(~SURVEY_YEAR)+
  theme_bw()
```
This is a big dataset with a lot of points. May be worth trying with sdmTMB, the issue is getting a CPUE from the hook counts.

Have a look at the observed catch per subsampling. The metadata say about the 'subsample' field: "Sampling scheme (1=20 hooks counted near the beginning of skate is standard in U.S. reg areas except one year in SEAK for yelloweye; 2=100% hook count is standard in Canadian reg areas)". However, the data contain 0 and 1 as values for SUBSAMPLE.
```{r}
subsample <- dat %>% select(OBS_CATCH,SUBSAMPLE,IPHC_REG)

t <- subsample %>% group_by(IPHC_REG,SUBSAMPLE) %>% tally() # area 2B (Canada) seems to be the only area where there is a prevalence of 0, which I'd interpret as (no subsampling)

subsample %>% #filter(OBS_CATCH>0) %>% 
  ggplot()+
  geom_boxplot(aes(x = factor(SUBSAMPLE), y=OBS_CATCH)) # 0 seems to present higher catches, hinting that 0 means no subsampling
```

It seems that if SUBSAMPLE=1 they do the subsampling, i.e. they only count the first 20 hooks. If SUBSAMPLING=0, they count all hooks on the skate. So, if SUBSAMPLING==1, OBS_CATCH=OBS_CATCH*5.
```{r}
sleep_all <- sleep_all %>% rowwise() %>% mutate(CATCH=ifelse(SUBSAMPLE==1,OBS_CATCH*5,OBS_CATCH))
```

Not 100% convinced that the adjustment for subsample is correct. Divide by hook number to get NPUE.
```{r}
sleep_all <- sleep_all %>% mutate(NPUE=CATCH/EX_EFF_HOOKS)
```

Prepare this for the sdmTMB code.
```{r}
sleep_all <- sleep_all %>% select(SURVEY_YEAR,FISHING_EVENT_ID,START_LAT,START_LON,AVG_DEPTH_M,NPUE) %>% 
  set_names(c(
  "year",
  "hauljoin",
  "lat",
  "lon",
  "depth",
  "npue"))

saveRDS(sleep_all,'../data/IPHC/sleep.RDS')
```

